Weighted Model Counting Without Parameter Variables

### 2021-06-17 doc/SAT_long_talk/talk.tex
Slides for the long SAT talk are almost finished

\begin{frame}[fragile]{The Problem of Computing Probability}
  \vspace{-0.75cm}
  \begin{columns}[t]
    \begin{column}{0.65\textwidth}
      \centering
      \begin{block}{ProbLog}
        \vspace{-0.2cm}
        \begin{lstlisting}[basicstyle=\tiny]
0.001 :: burglary.
0.002 :: earthquake.
0.95  :: alarm     :- burglary, earthquake.
0.94  :: alarm     :- burglary, \+ earthquake.
0.29  :: alarm     :- \+ burglary, earthquake.
0.001 :: alarm     :- \+ burglary, \+ earthquake.
0.9   :: johnCalls :- alarm.
0.05  :: johnCalls :- \+ alarm.
0.7   :: maryCalls :- alarm.
0.01  :: maryCalls :- \+ alarm.
        \end{lstlisting}
        \vspace{-0.2cm}
      \end{block}
      \vspace{-0.25cm}
      \begin{block}{BLOG}
        \vspace{-0.2cm}
        \begin{lstlisting}[escapeinside={(*}{*)},basicstyle=\tiny]
random Boolean Burglary (*$\sim$*) BooleanDistrib(0.001);
random Boolean Earthquake (*$\sim$*) BooleanDistrib(0.002);
random Boolean Alarm (*$\sim$*)
  if Burglary then
    if Earthquake then BooleanDistrib(0.95)
    else BooleanDistrib(0.94)
  else
    if Earthquake then BooleanDistrib(0.29)
    else BooleanDistrib(0.001);
random Boolean JohnCalls (*$\sim$*)
  if Alarm then BooleanDistrib(0.9)
  else BooleanDistrib(0.05);
random Boolean MaryCalls (*$\sim$*)
  if Alarm then BooleanDistrib(0.7)
  else BooleanDistrib(0.01);
        \end{lstlisting}
        \vspace{-0.2cm}
      \end{block}
    \end{column}
    \begin{column}{0.35\textwidth}
      \begin{block}{Bayesian Network}
        \centering
        \begin{tikzpicture}[node distance=2cm,scale=0.5,every node/.style={scale=0.5}]
          \node[draw,ellipse] (alarm) {Alarm};
          \node[draw,ellipse,above left of=alarm] (burglary) {Burglary};
          \node[draw,ellipse,above right of=alarm] (earthquake) {Earthquake};
          \node[draw,ellipse,below left of=alarm] (johnCalls) {JohnCalls};
          \node[draw,ellipse,below right of=alarm] (maryCalls) {MaryCalls};
          \draw[-Latex] (burglary) -- (alarm);
          \draw[-Latex] (earthquake) -- (alarm);
          \draw[-Latex] (alarm) -- (johnCalls);
          \draw[-Latex] (alarm) -- (maryCalls);
        \end{tikzpicture}
      \end{block}
      \vspace{1cm}
      \begin{block}{Markov Random Field}
        \centering
        \begin{tikzpicture}[node distance=2cm,scale=0.5,every node/.style={scale=0.5}]
          \node[draw,ellipse] (alarm) {Alarm};
          \node[draw,ellipse,above left of=alarm] (burglary) {Burglary};
          \node[draw,ellipse,above right of=alarm] (earthquake) {Earthquake};
          \node[draw,ellipse,below left of=alarm] (johnCalls) {JohnCalls};
          \node[draw,ellipse,below right of=alarm] (maryCalls) {MaryCalls};
          \draw (burglary) -- (earthquake);
          \draw (burglary) -- (alarm);
          \draw (earthquake) -- (alarm);
          \draw (alarm) -- (johnCalls);
          \draw (alarm) -- (maryCalls);
        \end{tikzpicture}
      \end{block}
    \end{column}
  \end{columns}
  \onslide<2>{
    \begin{tikzpicture}[remember picture,overlay]
      \node[draw,star,fill=red!10] (wmc) at (current page.center) {WMC};
      \coordinate[xshift=-0.25\linewidth,yshift=-0.25\textheight] (p1) at (current page.center);
      \coordinate[xshift=0.25\linewidth,yshift=-0.25\textheight] (p2) at (current page.center);
      \coordinate[xshift=-0.25\linewidth,yshift=0.25\textheight] (p3) at (current page.center);
      \coordinate[xshift=0.25\linewidth,yshift=0.25\textheight] (p4) at (current page.center);
      \draw[-latex,line width=2pt,color=red!50] (p1) -- (wmc);
      \draw[-latex,line width=2pt,color=red!50] (p2) -- (wmc);
      \draw[-latex,line width=2pt,color=red!50] (p3) -- (wmc);
      \draw[-latex,line width=2pt,color=red!50] (p4) -- (wmc);
    \end{tikzpicture}
  }
\end{frame}

@@

\section{An Alternative Formulation}
% TODO: Definition of WMC and MC-WMC

\begin{frame}{Formalising the Intuition from Before}
  For any propositional formula \structure{$\phi$} over a set of variables
  \structure{$X$} and \structure{$p, q \in \mathbb{R}$}, let
  \structure{$[\phi]^p_q\colon 2^X \to \mathbb{R}$} be the pseudo-Boolean
  function defined as
  \[
    [\phi]^p_q(Y) \coloneqq
    \begin{cases}
      p & \text{if } Y \models \phi \\
      q & \text{otherwise}
    \end{cases}
  \]
  for any \structure{$Y \subseteq X$}.

@@

\section{Correctness}

\section{Experimental Results}

\begin{frame}{Experimental Results}
  \centering
  \includegraphics[width=\textwidth]{cumulative}
\end{frame}

\begin{frame}{Experimental Results}
  \centering
  \includegraphics[width=\textwidth]{scatter1}
\end{frame}

\begin{frame}{Experimental Results}
  \centering
  \includegraphics[width=\textwidth]{scatter2}
\end{frame}

\begin{frame}{Experimental Results}
  \centering
  \includegraphics[width=\textwidth]{box}
\end{frame}
% TODO: fit the plot to the left half of the slide, add observations on the
% right

@@

\begin{frame}{Summary}
\end{frame}

>>>>

\begin{frame}[fragile]{The Computational Problem of Probabilistic Inference}
  \vspace{-0.75cm}
  \begin{columns}[t]
    \begin{column}{0.65\textwidth}
      \centering
      \begin{block}{ProbLog}
        \vspace{-0.2cm}
        \begin{lstlisting}[basicstyle=\tiny]
0.001 :: burglary.
0.002 :: earthquake.
0.95  :: alarm     :- burglary, earthquake.
0.94  :: alarm     :- burglary, \+ earthquake.
0.29  :: alarm     :- \+ burglary, earthquake.
0.001 :: alarm     :- \+ burglary, \+ earthquake.
0.9   :: johnCalls :- alarm.
0.05  :: johnCalls :- \+ alarm.
0.7   :: maryCalls :- alarm.
0.01  :: maryCalls :- \+ alarm.
        \end{lstlisting}
        \vspace{-0.2cm}
      \end{block}
      \vspace{-0.25cm}
      \begin{block}{BLOG}
        \vspace{-0.2cm}
        \begin{lstlisting}[escapeinside={(*}{*)},basicstyle=\tiny]
random Boolean Burglary (*$\sim$*) BooleanDistrib(0.001);
random Boolean Earthquake (*$\sim$*) BooleanDistrib(0.002);
random Boolean Alarm (*$\sim$*)
  if Burglary then
    if Earthquake then BooleanDistrib(0.95)
    else BooleanDistrib(0.94)
  else
    if Earthquake then BooleanDistrib(0.29)
    else BooleanDistrib(0.001);
random Boolean JohnCalls (*$\sim$*)
  if Alarm then BooleanDistrib(0.9)
  else BooleanDistrib(0.05);
random Boolean MaryCalls (*$\sim$*)
  if Alarm then BooleanDistrib(0.7)
  else BooleanDistrib(0.01);
        \end{lstlisting}
        \vspace{-0.2cm}
      \end{block}
    \end{column}
    \begin{column}{0.35\textwidth}
      \begin{block}{Bayesian Network}
        \centering
        \begin{tikzpicture}[node distance=2cm,scale=0.5,every node/.style={scale=0.5}]
          \node[draw,ellipse] (alarm) {Alarm};
          \node[draw,ellipse,above left of=alarm] (burglary) {Burglary};
          \node[draw,ellipse,above right of=alarm] (earthquake) {Earthquake};
          \node[draw,ellipse,below left of=alarm] (johnCalls) {JohnCalls};
          \node[draw,ellipse,below right of=alarm] (maryCalls) {MaryCalls};
          \draw[-Latex] (burglary) -- (alarm);
          \draw[-Latex] (earthquake) -- (alarm);
          \draw[-Latex] (alarm) -- (johnCalls);
          \draw[-Latex] (alarm) -- (maryCalls);
        \end{tikzpicture}
      \end{block}
      \vspace{1cm}
      \begin{block}{Markov Random Field}
        \centering
        \begin{tikzpicture}[node distance=2cm,scale=0.5,every node/.style={scale=0.5}]
          \node[draw,ellipse] (alarm) {Alarm};
          \node[draw,ellipse,above left of=alarm] (burglary) {Burglary};
          \node[draw,ellipse,above right of=alarm] (earthquake) {Earthquake};
          \node[draw,ellipse,below left of=alarm] (johnCalls) {JohnCalls};
          \node[draw,ellipse,below right of=alarm] (maryCalls) {MaryCalls};
          \draw (burglary) -- (earthquake);
          \draw (burglary) -- (alarm);
          \draw (earthquake) -- (alarm);
          \draw (alarm) -- (johnCalls);
          \draw (alarm) -- (maryCalls);
        \end{tikzpicture}
      \end{block}
    \end{column}
  \end{columns}
  \onslide<2>{
    \begin{tikzpicture}[remember picture,overlay]
      \node[draw,star,fill=red!10] (wmc) at (current page.center) {WMC};
      \coordinate[xshift=-0.25\linewidth,yshift=-0.25\textheight] (p1) at (current page.center);
      \coordinate[xshift=0.25\linewidth,yshift=-0.25\textheight] (p2) at (current page.center);
      \coordinate[xshift=-0.25\linewidth,yshift=0.25\textheight] (p3) at (current page.center);
      \coordinate[xshift=0.25\linewidth,yshift=0.25\textheight] (p4) at (current page.center);
      \draw[-latex,line width=2pt,color=red!50] (p1) -- (wmc);
      \draw[-latex,line width=2pt,color=red!50] (p2) -- (wmc);
      \draw[-latex,line width=2pt,color=red!50] (p3) -- (wmc);
      \draw[-latex,line width=2pt,color=red!50] (p4) -- (wmc);
    \end{tikzpicture}
  }
\end{frame}

@@

\section{A More Expressive Alternative}

\begin{frame}{WMC, Formally}
  \begin{definition}
    A \alert{WMC instance} is a tuple \structure{$(\phi, X_I, X_P, w)$}, where
    \begin{itemize}
    \item \structure{$X_I$} is the set of \alert{indicator variables},
    \item \structure{$X_P$} is the set of \alert{parameter variables} (with
      \structure{$X_I \cap X_P = \emptyset$}),
    \item \structure{$\phi$} is a propositional formula in CNF over
      \structure{$X_I \cup X_P$},
    \item \structure{$w\colon X_I \cup X_P \cup \{\neg x \mid x \in X_I \cup
        X_P\} \to \mathbb{R}$} is a \alert{weight function}
      \begin{itemize}
      \item such that \structure{$w(x) = w(\neg x) = 1$} for all \structure{$x
          \in X_I$}.
      \end{itemize}
    \end{itemize}
  \end{definition}
  \begin{definition}
    Let \structure{$\phi$} be a formula over a set of variables \structure{$X$}.
    Then \structure{$Y \subseteq X$} is a \alert{minimum-cardinality model} of
    \structure{$\phi$} if
    \begin{itemize}
    \item \structure{$Y \models \phi$},
    \item and \structure{$|Y| \le |Z|$} for all \structure{$Z \models \phi$}.
    \end{itemize}
  \end{definition}
\end{frame}

\begin{frame}{WMC and Minimum-Cardinality WMC}
  The goal of \alert{WMC} is to compute
  \[
    \sum_{Y \models \phi} \prod_{Y \models l} w(l)
  \]
  whereas the goal of \alert{minimum-cardinality WMC} is to compute
  \[
    \sum_{Y \models \phi\text{, }\alert{|Y| = k}} \prod_{Y \models l} w(l),
  \]
  where
  \[
    k = \min_{Y \models \phi} |Y|.
  \]
\end{frame}

\begin{frame}{A More Expressive Alternative}
  For any propositional formula \structure{$\phi$} over a set of variables
  \structure{$X$} and \structure{$p, q \in \mathbb{R}$}, let
  \structure{$[\phi]^p_q\colon 2^X \to \mathbb{R}$} be the pseudo-Boolean
  function defined as
  \[
    [\phi]^p_q(Y) \coloneqq
    \begin{cases}
      p & \text{if } Y \models \phi \\
      q & \text{otherwise}
    \end{cases}
  \]
  for any \structure{$Y \subseteq X$}.

@@

\section{When Does This Transformation Work?}

\begin{frame}{Correctness Conditions for WMC (1/2)}
  For each parameter variable \structure{$p \in X_P$},
  \begin{itemize}
  \item \structure{$w(\neg p) = 1$},
  \item and the set of clauses that mention \structure{$p$} or \structure{$\neg
      p$} is
    \[
      \left\{p \lor \bigvee_{i=1}^n \neg l_i \right\} \cup \{ l_i \lor \neg p
      \mid i = 1, \dots, n \}
    \]
    for some non-empty family of \alert{indicator} literals
    \structure{$(l_i)_{i=1}^n$}.
  \end{itemize}
\end{frame}

\begin{frame}{Correctness Conditions for WMC (2/2)}
  For each parameter variable \structure{$p \in X_P$},
  \begin{itemize}
  \item \structure{$w(p) + w(\neg p) = 1$},
  \item each clause has at most one parameter variable,
  \item there is no clause \structure{$c \in \phi$} such that \structure{$\neg p
      \in c$},
  \item if \structure{$\{ p \} \in \phi$}, then this is the only clause that
    mentions \structure{$p$},
  \item and for any \structure{$c, d \in \phi$} such that \structure{$c \ne d$},
    \structure{$p \in c$} and \structure{$p \in d$},
    \[
      \bigwedge_{l \in c \setminus \{ p \}} \neg l \land \bigwedge_{l \in d
        \setminus \{ p \}} \neg l
    \]
    is false.
  \end{itemize}
\end{frame}

\begin{frame}{Additional Conditions for Minimum-Cardinality WMC}
  \begin{itemize}
  \item All models of \structure{$\{ c \in \phi \mid c \cap X_P = \emptyset \}$}
    have the same number of positive indicator literals,
  \item and
    \[
      \min_{Z \subseteq X_P} |Z| \quad \text{s.t. } \quad Y \cup Z \models \phi
    \]
    is the same for all \structure{$Y \models \{ c \in \phi \mid c \cap X_P =
      \emptyset \}$}.
  \end{itemize}
\end{frame}

\section{How Good Is It?}

\begin{frame}{WMC/PBP Encodings for Bayesian Networks}
  \centering
  \includegraphics[width=\textwidth]{cumulative1}
\end{frame}

\begin{frame}{WMC/PBP Encodings for Bayesian Networks}
  \centering
  \includegraphics[width=\textwidth]{cumulative2}
\end{frame}

\begin{frame}{WMC/PBP Encodings for Bayesian Networks}
  \centering
  \includegraphics[width=\textwidth]{cumulative3}
\end{frame}

\begin{frame}{WMC/PBP Encodings for Bayesian Networks}
  \centering
  \includegraphics[width=\textwidth]{cumulative4}
\end{frame}

\begin{frame}{WMC/PBP Encodings for Bayesian Networks}
  \centering
  \includegraphics[width=\textwidth]{cumulative5}
\end{frame}

\begin{frame}{Compared to the Previous State of the Art}
  \centering
  \includegraphics[width=\textwidth]{scatter1}
\end{frame}

\begin{frame}{The Best Encoding for DPMC: Before and After}
  \centering
  \includegraphics[width=\textwidth]{scatter2}
\end{frame}

% \begin{frame}{Variable Reduction for Bayesian Network Encodings}
%   \begin{columns}
%     \begin{column}{0.4\textwidth}
%       \begin{itemize}
%       \item foo
%       \end{itemize}
%     \end{column}
%     \begin{column}{0.6\textwidth}
%       \centering
%       \includegraphics[width=\textwidth]{box}
%     \end{column}
%   \end{columns}
% \end{frame}

@@

\begin{frame}{Summary}
  \begin{itemize}
  \item TODO
  \end{itemize}
\end{frame}
